---
title: "MiDAS_course"
author: "Miriam Peces, Sofie Zacho Vestergaard, Marta Nierychlo"
date: "`r format(Sys.time(), '%Y-%m-%d')`, Aalborg, Denmark"
output: html_document
---

# Questions for the MiDAS course - ampvis hands on. Whenever you encounter "..." it means you should fill something in there.

## Only do ONCE!!!! 
```{r, eval = F}
install.packages("remotes")
remotes::install_github("kasperskytte/ampvis2")
install.packages("tidyverse")
```


## Load the packages
```{r, warning = F, message=F}
library(ampvis2)
library(tidyverse)
```


## Set your working directory - only if you did not open script from the downloaded folder
```{r, eval = FALSE}
#setwd("") 
```


## Load the data into an ampvis object
```{r}
d_midas <- amp_load(otutable = "data/ASVtable.tsv",
              taxonomy = "data/ASVs.R1.midas53.sintax",
              metadata = "data/metadata_midascourse.txt") 
```


# Inspect the data - get a summary of the ampvis object
```{r}
d_midas
```

# Investigate rarefaction curve
```{r}
amp_rarecurve(d_midas, 
              color_by = 'SampleSite')
```

## plot the number of reads for each sample
```{r}
ggplot(amp_alpha_diversity(d_midas))+
  geom_point(aes(x= SampleID, y = Reads, color=SampleSite)) +
  facet_grid(~ SampleSite, scales = 'free_x', space = 'free') +
  theme(axis.text.x = element_text(angle = 90, size=6, vjust= 0.5))

```

# Filter data: Remove low read samples and normalise data
```{r}
dsn_midas <- amp_subset_samples(d_midas,
                                minreads = 10000, # remove samples with less than 10000 reads
                                normalise = TRUE)
```


Questions to answer

Q1 What are the top 25 most abundant genera?
	hint: look into amp_heatmap()
```{r}
amp_heatmap(dsn_midas,
            normalise = F,
            group_by = ...,
            tax_show = ...,
            tax_aggregate = ...,
            ...)
```

	
Q2 What are the top 25 most abundant species? And what genera are these belonging to (plot these with the species)?
  Addition (optional) try to sort based on one of the plants
```{r}
amp_heatmap(dsn_midas,
            normalise = F,
            group_by = ...,
            tax_show = ...,
            tax_aggregate = ...,
            tax_add = ...,
            ...)
```


Q3 Try to save the heatmap from before
```{r}
heatmap <- amp_heatmap(dsn_midas,
            normalise = F,
            group_by = ...,
            tax_show = ...,
            tax_aggregate = ...,
            tax_add = ...,
            ...)

ggsave("heatmap.png", heatmap,
       width = 8, height = 6 #change to fit your needs
       ) 

```


Q4. Perform an ordination to compare the microbial structures of the different WWTPs. 
```{r}
amp_ordinate(data = dsn_midas, 
             transform = "Hellinger", 
             type = "PCA",
             sample_color_by = "SampleSite")
```



Compare to silva taxonomy
Q5 How many of the ASVs are classified at genus and species level using silva?
	hint: amp_load() with silva tax
```{r}
d_silva <- amp_load(otutable = "data/ASVtable.tsv",
              taxonomy = "data/ASVs_SILVA_138.2.R1.sintax",
              metadata = "data/metadata_midascourse.txt") 

dsn_silva <- amp_subset_samples(d_silva,
                                minreads = 10000, # remove samples with less than 10000 reads
                                normalise = TRUE)

dsn_silva

```
	
	
Q6 What are the most abundant 25 genera using silva? Compare to MiDAS taxonomic assignment
```{r}
amp_heatmap(dsn_silva,
            normalise = F,
            group_by = ...,
            tax_show = ...,
            tax_aggregate = ...,
            ...)
```
	



Q7 Make a timeseries plot for Ca. Phosphoribacter and Ca. Accumulibacter in WWTP_C?
	hint: look into amp_subset_taxa(), amp_subset_sample(), amp_time_series()
```{r}
subset <- amp_subset_taxa(data = dsn_midas, tax_vector = c("g__Ca_Phosphoribacter", "g__Ca_Accumulibacter")) %>% 
  amp_subset_samples(SampleSite == "WWTP_C")

subset$metadata$SampleDate <- as.Date(subset$metadata$SampleDate, format = c("%d/%m/%Y"))
  
amp_time_series(subset,
                time_variable = "SampleDate", 
                tax_aggregate = "Genus",
                normalise = FALSE)
```

	
Q8 Make a timeseries plot for a species and plant of your choice
```{r}
subset <- amp_subset_taxa(data = dsn_midas, tax_vector = c("...")) %>% 
  amp_subset_samples(SampleSite == "...")

subset$metadata$SampleDate <- as.Date(subset$metadata$SampleDate, format = c("%d/%m/%Y"))
  
amp_time_series(subset,
                time_variable = "SampleDate", 
                tax_aggregate = "Genus",
                normalise = FALSE)
```


# extra questions 
Qe1-a Make an ordination plot for all plants - Try different ordinations, e.g PCoA based on bray-curtis distance
  hint: look into amp_ordinate()
```{r}
amp_ordinate(dsn_midas, 
             sample_color_by = '...',
             ...)

```

Qe1-b Make an ordination plot for a plant of your choice to explore the temporal dynamics of the overall microbial structure
  hint: subset first for your chosen WWTP
```{r}
plant_subset <- amp_subset_samples(SampleSite == "...")

amp_ordinate(plant_subset, 
             sample_trajectory = "...", 
             sample_color_by = "...",
             sample_label_by = "...",
             ...)

```


Qe2 How many of the top 25 genera are filamentous?
	hint: look into amp_heatmap(…, plot_functions = TRUE). Needs tax_aggregate at genus-level
```{r}
amp_heatmap(dsn_midas, 
             ...)

```
	
Qe3 Which genera within Proteobacteria are the most abundant across all samples?
	hint: look into amp_subset_taxa()
```{r}

```
	
Qe4 Make alpha diversity analysis for the data (remember to rarefy at the same time)
  hint: look into amp_alpha_diversity()
```{r}

```








