---
title: "MiDAS_course_2025"
author: "Miriam Peces, Sofie Zacho Vestergaard, Marta Nierychlo"
date: "`r format(Sys.time(), '%Y-%m-%d')`, Aalborg, Denmark"
output: html_document
---


## Only do ONCE!!!! 
```{r, eval = F}
install.packages("remotes")
remotes::install_github("kasperskytte/ampvis2")
install.packages("tidyverse")
```


## Load the packages
```{r, warning = F, message=F}
library(ampvis2)
library(tidyverse)
```


## Set your working directory - only if you did not open script from the downloaded folder
```{r, eval = FALSE}
#setwd("") 
```


## Load the data into an ampvis object
```{r}
d_midas <- amp_load(otutable = "data/ASVtable.tsv",
              taxonomy = "data/ASVs.R1.midas53.sintax",
              metadata = "data/metadata_midascourse.txt",
              fasta = "data/ASVs.R1.fa") # This one is optional, but handy if you need to merge datasets processed in different runs
```


# Inspect the data - get a summary of the ampvis object
```{r}
d_midas
```

# Investigate rarefaction curve
```{r}
amp_rarecurve(d_midas, 
              color_by = 'SampleSite') +
  geom_vline(xintercept = 10000)
```

## plot the number of reads for each sample
```{r}

ggplot( amp_alpha_diversity(d_midas))+
  geom_point(aes(x= SampleID, y = Reads, color=SampleSite)) +
  facet_grid(~ SampleSite, scales = 'free_x', space = 'free') +
  theme(axis.text.x = element_text(angle = 90, size=6, vjust= 0.5))

```



# Filter data: Remove low read samples and normalise data
```{r}
dsn_midas <- amp_subset_samples(d_midas,
                                minreads = 10000, # remove samples with less than 10000 reads
                                normalise = TRUE)
```





Questions to answer

Q1 What are the top 25 most abundant genera?
	hint: look into amp_heatmap()
```{r}
amp_heatmap(dsn_midas,
            normalise = F,
            group_by = 'SampleSite',
            tax_show = 25,
            tax_aggregate = 'Genus',
            color_vector = c('white', 'darkred'))
```

	
Q2 What are the top 25 most abundant species? And what genera are these belonging to (plot these with the species)?
```{r}
amp_heatmap(dsn_midas,
            normalise = F,
            group_by = 'SampleSite',
            tax_show = 25,
            tax_aggregate = 'Species',
            tax_add = 'Genus',
            color_vector = c('white', 'darkred'))
```
Addition Q3 - sort by a specific plant
```{r}
amp_heatmap(dsn_midas,
            normalise = F,
            group_by = 'SampleSite',
            tax_show = 25,
            tax_aggregate = 'Species',
            sort_by = 'Randers',
            tax_add = 'Genus',
            color_vector = c('white', 'darkred'))
```

Q3 Try to save the heatmap from before
```{r}
heatmap <- amp_heatmap(dsn_midas,
            normalise = F,
            group_by = 'SampleSite',
            tax_show = 25,
            tax_aggregate = 'Species',
            tax_add = 'Genus',
            color_vector = c('white', 'darkred'))

ggsave("heatmap.png", heatmap,
       width = 8, height = 6 #change to fit your needs
       ) 

```





Compare to silva taxonomy
Q4 How many of the ASVs are classified at genus and species level using silva?
	hint: amp_load() with silva tax
```{r}
d_silva <- amp_load(otutable = "data/ASVtable.tsv",
              taxonomy = "data/ASVs_SILVA_138.2.R1.sintax",
              metadata = "data/metadata_midascourse.txt",
              fasta = "data/ASVs.R1.fa") 

dsn_silva <- amp_subset_samples(d_silva,
                                minreads = 10000, # remove samples with less than 10000 reads
                                normalise = TRUE)

dsn_silva

```
	
	
Q5 What are the most abundant 25 genera using silva? Compare to MiDAS taxonomic assignment
```{r}
amp_heatmap(dsn_silva,
            normalise = F,
            group_by = 'SampleSite',
            tax_show = 25,
            tax_aggregate = 'Species',
            tax_add = 'Genus',
            color_vector = c('white', 'darkred'))
```
	

Q6 Make a timeseries plot for Ca. Phosphoribacter and Ca. Accumulibacter in Randers?
	hint: look into amp_subset_taxa(), amp_subset_sample(), amp_time_series()
```{r}
subset <- amp_subset_taxa(data = dsn_midas, tax_vector = c("g__Ca_Phosphoribacter", "g__Ca_Accumulibacter")) %>% 
  amp_subset_samples(SampleSite == "Randers")

subset$metadata$SampleDate <- as.Date(subset$metadata$SampleDate, format = c("%d/%m/%Y"))
  
amp_time_series(subset,
                time_variable = "SampleDate", 
                tax_aggregate = "Genus",
                normalise = FALSE)
```

	
Q7 Make a timeseries plot for a species and plant of your choice
```{r}
subset <- amp_subset_taxa(data = dsn_midas, tax_vector = c("g__Zoogloea")) %>% 
  amp_subset_samples(SampleSite == "Randers")

subset$metadata$SampleDate <- as.Date(subset$metadata$SampleDate, format = c("%d/%m/%Y"))
  
amp_time_series(subset,
                time_variable = "SampleDate", 
                tax_aggregate = "Genus",
                normalise = FALSE)
```




# extract questions 
Qe1 Make an ordination plot for all plants - Try different ordinations, e.g PCoA based on bray-curtis distance
  hint: look into amp_ordinate()
```{r}
amp_ordinate(dsn_midas, 
             sample_color_by = 'SampleSite')

amp_ordinate(dsn_midas, 
             sample_color_by = 'SampleSite',
             type = 'PCoA',
             distmeasure = 'Bray',
             sample_colorframe = T,
             sample_colorframe_label = 'SampleSite')

```

Qe2 How many of the top 25 genera are filamentous?
	hint: look into amp_heatmap(…, plot_functions = TRUE). Needs tax_aggregate at genus-level
```{r}
amp_heatmap(dsn_midas, 
             normalise = F,
            tax_aggregate = 'Genus',
            plot_functions = T,
            group_by = 'SampleSite',
            tax_show = 25,
            sort_by = 'Lynetten')

```
	
Qe3 Which genera within Proteobacteria are the most abundant across all samples?
	hint: look into amp_subset_taxa()
```{r}

proteo <- amp_subset_taxa(dsn_midas,
                          tax_vector = 'p__Proteobacteria',
                          normalise = F)

amp_heatmap(proteo, 
             normalise = F,
            tax_aggregate = 'Genus',
            group_by = 'SampleSite',
            tax_show = 25)

```
	
Qe4 Make alpha diversity analysis for the data (remember to rarefy at the same time)
  hint: look into amp_alpha_diversity()
```{r}

alpha <- amp_alpha_diversity(amp_subset_samples(d_midas,
                                minreads = 10000, normalise = F),
                            rarefy = 10000)


ggplot(alpha) +
  geom_boxplot(aes(x=SampleSite, y = ObservedOTUs, fill = SampleSite),
              )

ggplot(alpha) +
  geom_boxplot(aes(x=SampleSite, y = Shannon, fill = SampleSite)) +
  geom_point(aes(x=SampleSite, y = Shannon, fill = SampleSite),
             shape=21, position = position_jitter(width = 0.1))

```








