---
title: "MiDAS_course_2025"
author: "Miriam Peces, Sofie Zacho Vestergaard, Marta Nierychlo"
date: "`r format(Sys.time(), '%Y-%m-%d')`, Aalborg, Denmark"
output: html_document
---


```{r, eval = F}
#install.packages("remotes")
#remotes::install_github("kasperskytte/ampvis2")
```


## Load the packages
Remember to load the packages before you try to use the associated functions.
```{r, warning = F, message=F}
library(ampvis2)
library(tidyverse)
library(data.table)
```

# **1. Load and prepare the data for analysis** 
## 1.1 Set your working directory
```{r, eval = FALSE}
setwd("..change to your directory..") 
```


## 1.2 Load the data into R
```{r}
d_midas <- amp_load(otutable = "data/ASVtable.tsv",
              taxonomy = "data/ASVs.R1.midas53.sintax",
              metadata = "data/metadata_midascourse.txt",
              fasta = "data/ASVs.R1.fa") # This one is optional, but handy if you need to merge datasets processed in different runs
```

## 1.4 Add/modify metadata
```{r}
# Check the data types in the metadata
str(d_midas$metadata)

# Let's modify the type of SampleDate
d_midas$metadata$SampleDate <- as.Date(d_midas$metadata$SampleDate, format = c("%d/%m/%Y"))


# Add month information as character
d_midas$metadata$Month <- as.character(format(as.Date(d_midas$metadata$SampleDate),'%m'))

# Tip, use %B if you prefer to use the month name, and use factor() to order the month names chronologically
d_midas$metadata$MonthName <- factor(format(as.Date(d_midas$metadata$SampleDate),'%B'),
                                     levels = month.name)
```

# Normalise data
```{r}
dsn_midas <- amp_subset_samples(d_midas,
                                minreads = 10000,
                                SampleID != "MQ201118-182",  
                                normalise = TRUE)
```




Questions to answer

Most abundant bacteria
Q1 What is the top 25 most abundant genera?
	hint: look into amp_heatmap()
Q2 What is the top 25 most abundant species? And what genera are these belonging to (plot these with the species)?
Q3 What is the top 25 most abundant species in Ribe?
Q4 How many of the top 25 genera are filamentous and important for phosphorus removal?
	hint: look into amp_heatmap(…, plot_functions = TRUE)
Q5 Which genera within Proteobacteria are the most abundant across all samples?
	hint: look into amp_subset_taxa()

Compare to silva taxonomy
Q6 How many of the ASVs are classified at genus and species level using silva?
	hint: amp_load() with silva tax
Q7 What are the most abundant 25 genera using silva?
	

Timeseries
Q8 Can you make a timeserie plot for Ca. Phosphoribacter and Ca. Accumulibacter in Randers?
	hint: look into amp_subset_taxa(), amp_subset_sample(), amp_time_series()
Q9 Make a timeseries plot for a species and plant of your chouce
Q10 Make a timeserie ordination for all the plants


Core communities 
Q11 What is the strict and general core species? (0.1% relative abundance in 80% (strict) and 50% (general) of the plants)
	hint: Look into last part of the .Rmd file with answers you looked at yesterday






