---
title: "MiDAS_course_2025"
author: "Miriam Peces, Sofie Zacho Vestergaard, Marta Nierychlo"
date: "`r format(Sys.time(), '%Y-%m-%d')`, Aalborg, Denmark"
output: html_document
---


# Only do ONCE
```{r, eval = F}
install.packages("remotes")
remotes::install_github("kasperskytte/ampvis2")
install.packages("tidyverse")
```


## Load the packages
Remember to load the packages before you try to use the associated functions.
```{r, warning = F, message=F}
library(ampvis2)
library(tidyverse)
```


# **1. Load and prepare the data for analysis** 
## 1.1 Set your working directory - only if you did not open script from the downloaded folder
```{r, eval = FALSE}
#setwd("") 
```


## 1.2 Load the data into R
```{r}
d_midas <- amp_load(otutable = "data/ASVtable.tsv",
              taxonomy = "data/ASVs.R1.midas53.sintax",
              metadata = "data/metadata_midascourse.txt",
              fasta = "data/ASVs.R1.fa") # This one is optional, but handy if you need to merge datasets processed in different runs
```


# Inspect the data - get a summary of the ampvis object
```{r}
d_midas
```

# Investigate rarefaction curve
```{r}
amp_rarecurve(d_midas, 
              color_by = 'SampleSite')
```

## plot the number of reads for each sample
```{r}

ggplot( amp_alpha_diversity(d_midas))+
  geom_point(aes(x= SampleID, y = Reads, color=SampleSite)) +
  facet_grid(~ SampleSite, scales = 'free_x', space = 'free') +
  theme(axis.text.x = element_text(angle = 90, size=6, vjust= 0.5))

```



# Filter data: Remove low read samples and normalise data
```{r}
dsn_midas <- amp_subset_samples(d_midas,
                                minreads = 10000, # remove samples with less than 10000 reads
                                normalise = TRUE)
```





Questions to answer

Q1 What are the top 25 most abundant genera?
	hint: look into amp_heatmap()
```{r}
amp_heatmap(dsn_midas,
            normalise = F,
            group_by = ...,
            tax_show = ...,
            tax_aggregate = ...,
            ...)
```

	
Q2 What are the top 25 most abundant species? And what genera are these belonging to (plot these with the species)?
```{r}
amp_heatmap(dsn_midas,
            normalise = F,
            group_by = ...,
            tax_show = ...,
            tax_aggregate = ...,
            tax_add = ...,
            ...)
```


Q3 Try to save the heatmap from before
```{r}
heatmap <- amp_heatmap(dsn_midas,
            normalise = F,
            group_by = ...,
            tax_show = ...,
            tax_aggregate = ...,
            tax_add = ...,
            ...)

ggsave("heatmap.png", heatmap,
       width = 8, height = 6 #change to fit your needs
       ) 

```





Compare to silva taxonomy
Q4 How many of the ASVs are classified at genus and species level using silva?
	hint: amp_load() with silva tax
```{r}
d_silva <- amp_load(otutable = "data/ASVtable.tsv",
              taxonomy = "data/ASVs_SILVA_138.2.R1.sintax",
              metadata = "data/metadata_midascourse.txt",
              fasta = "data/ASVs.R1.fa") 

dsn_silva <- amp_subset_samples(d_silva,
                                minreads = 10000, # remove samples with less than 10000 reads
                                normalise = TRUE)

dsn_silva

```
	
	
Q5 What are the most abundant 25 genera using silva? Compare to MiDAS taxonomic assignment
```{r}
amp_heatmap(dsn_silva,
            normalise = F,
            group_by = ...,
            tax_show = ...,
            tax_aggregate = ...,
            ...)
```
	

Q6 Make a timeseries plot for Ca. Phosphoribacter and Ca. Accumulibacter in Randers?
	hint: look into amp_subset_taxa(), amp_subset_sample(), amp_time_series()
```{r}
subset <- amp_subset_taxa(data = dsn_midas, tax_vector = c("g__Ca_Phosphoribacter", "g__Ca_Accumulibacter")) %>% 
  amp_subset_samples(SampleSite == "Randers")

subset$metadata$SampleDate <- as.Date(subset$metadata$SampleDate, format = c("%d/%m/%Y"))
  
amp_time_series(subset,
                time_variable = "SampleDate", 
                tax_aggregate = "Genus",
                normalise = FALSE)
```

	
Q7 Make a timeseries plot for a species and plant of your choice
```{r}
subset <- amp_subset_taxa(data = dsn_midas, tax_vector = c("...")) %>% 
  amp_subset_samples(SampleSite == "...")

subset$metadata$SampleDate <- as.Date(subset$metadata$SampleDate, format = c("%d/%m/%Y"))
  
amp_time_series(subset,
                time_variable = "SampleDate", 
                tax_aggregate = "Genus",
                normalise = FALSE)
```




# extract questions 
Qe1 Make an ordination plot for all plants
  hint: look into amp_ordinate()
Qe2 How many of the top 25 genera are filamentous?
	hint: look into amp_heatmap(…, plot_functions = TRUE). Needs tax_aggregate at genus-level
Qe3 Which genera within Proteobacteria are the most abundant across all samples?
	hint: look into amp_subset_taxa()








