---
title: "MiDAS course: Hands on - part 2"
author: "Marie Rissgaard-Jensen"
date: "`r format(Sys.time(), '%Y-%m-%d')`, Aalborg, Denmark"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Questions for the MiDAS course - ampvis hands on part 2. 


### Load the packages

```{r, warning = F, message=F}
library(ampvis2)
library(tidyverse)
library(patchwork) ## Combine different plot using '+'
library(vegan) ## To calculate distance matrix between samples  

```

### Set your working directory - only if you did not open script from the downloaded folder
```{r}
setwd("C:/Users/HD95LP/Downloads/output/MiDAS course - hands on part 2") 
WD = "C:/Users/HD95LP/Downloads/output/MiDAS course - hands on part 2/"

```


# Details about the pipelines


**ASV based: **
Pipeline: https://github.com/KasperSkytte/snakemake_usearch 



**Alignment based (95% identity)**
Pipeline: https://github.com/cmc-aau/nanopore_16Samp 



## Load the data into an ampvis object

- Remember to adjust paths

```{r}

metadata <- read.csv(paste0(WD, "metadata.csv")) %>%
  mutate(SampleID = Barcode) %>% relocate(SampleID) %>% 
  mutate(PlantID = case_when(LibID == "PCR pos" ~ "PCR pos",
                             LibID == "PCR neg" ~ "PCR neg", .default = PlantID
                             ))

# Alignment based (minimap min identity 95%)
## Raw read (in counts)
d_alignment_raw <- amp_load(otutable = paste0(WD, "Alignment-based/alignment_otutable_mappedreads.tsv"),
              metadata = metadata) 
## Normalised data (in percentage: with relative abundances are calculated taking non-mapped reads into account)
d_alignment_norm <- amp_load(otutable = paste0(WD, "Alignment-based/alignment_otutable_normalised.tsv"),
              metadata = metadata) 

## Summary file 
summary_reads_align <- vroom::vroom(paste0(WD, "Alignment-based/alignment_summary.txt"), show_col_types = F) 

# ASV based 
d_ASV <- amp_load(otutable = paste0(WD, "ASV-based/ASV_abund_table.tsv"),
              taxonomy = paste0(WD, "ASV-based/ASV_zOTUs.sintax"),
              metadata = metadata
              ) 
# Total reads 
summary_reads_ASV <- read.csv(paste0(WD, "ASV-based/ASV_totalreads.csv"), header = F, col.names = c("SampleID", "total_reads")) 

```

# QC 


### Ordination 
```{r}

amp_ordinate(d_ASV, sample_color_by = "PlantID") + 
        labs(subtitle = "ASV-based") +
  amp_ordinate(d_alignment_norm, sample_color_by = "PlantID") +
          labs(subtitle = "alignment-based")  + 
  plot_layout(guides = "collect")

```

### Ordination - no control

```{r}
amp_ordinate(d_ASV %>% amp_subset_samples(!str_detect(PlantID,  "PCR")), sample_color_by = "PlantID") + 
        labs(subtitle = "ASV-based") +
  amp_ordinate(d_alignment_norm %>% amp_subset_samples(!str_detect(PlantID,  "PCR")), sample_color_by = "PlantID") +
          labs(subtitle = "alignment-based") + 
  plot_layout(guides = "collect")

```


### Rarefaction 

```{r}
amp_rarecurve(d_ASV, color_by =  "PlantID") + 
      labs(subtitle = "ASV-based") +
  amp_rarecurve(d_alignment_raw, color_by = "PlantID") +
        labs(subtitle = "alignment-based")  + 
  plot_layout(guides = "collect")
```



# Questions: 

Explain and/or make plots in order to answer the questions

- What are the main difference between the datasets? 
- How are unclassified reads treated?
- How many reads are unmapped in the two datasets? What do these represent? 
- Calculate alpha diversity indexes - what do they tell. What is the importance of rarefaction 
- How many distinct ASVs/Species/Genus/.. are present for each dataset - how many are shared? What is the effect of adding an abundance cutoff? 
- What are the most abundant taxa - does if differ depending on taxonomic level?
- Consider how to merge dataset to visualize betadiversity
- At which taxonomic level would you trust the datasets? Why?


# How are unclassified reads treated?

**ASV-based**: Novelty is clustered into ASVs without any classification


**Alignment-based**: No match: the read will be put into a single unclassified group


### Unclassified when using an ASV based approch 

```{r}

d_ASV$tax %>% 
  mutate(Order = if_else(Order == "", " Unclassified", Order)) %>% 
  arrange((Order)) %>% head()


d_ASV$tax %>% 
  mutate(Family = if_else(Family == "", " Unclassified", Order)) %>% 
  arrange((Family)) %>% head(n = 10)


d_ASV$tax %>% 
  mutate(Genus = if_else(Genus == "", " Unclassified", Order)) %>% 
  arrange((Genus)) %>% head(n = 10)


```


### Summary of read/abundance in the ampvis object

```{r}


d_alignment_norm$abund %>% colSums()
d_alignment_raw$abund %>% colSums()

## ASV based
d_ASV$abund %>% colSums()
amp_subset_samples(d_ASV, normalise = T)[[1]] %>% colSums()

```



# How many reads are unmapped in the two datasets? What do these represent?

```{r}
merged_alpha_div <- 
  rbind(
    d_alignment_raw %>% amp_alpha_diversity() %>% mutate(type = "alignment") %>%
      left_join(summary_reads_align %>% select(barcode, total_filtered_reads) %>% 
                  rename("SampleID" = "barcode", 
                         "total_reads" = "total_filtered_reads"), by = join_by(SampleID)),
    d_ASV %>% amp_alpha_diversity() %>% mutate(type = "ASV") %>%
      left_join(summary_reads_ASV, by = join_by(SampleID)))
```

### Mapped read summery

```{r}
merged_alpha_div %>% 
ggplot(
    aes(x = PlantID, y = Reads, colour = type)
  ) + 
    geom_point(aes(group = type)) + 
  geom_boxplot() 

```

### Total and mapped reads summery
```{r}
merged_alpha_div %>% 
    pivot_longer(cols = c(total_reads, Reads)) %>% 
  ggplot(
    aes(x = type, y = value, colour = name)
  ) + 
    geom_point() + 
  geom_boxplot() +
    #labs(subtitle = "Rarefied 70137 reads") +
    facet_wrap(~PlantID, scales = "free_y", nrow = 2)
```

### Percent mapped reads 
```{r}
merged_alpha_div %>% 
    filter(!str_detect(PlantID,  "PCR")) %>% 
  mutate(p = Reads/total_reads*100) %>% 
    pivot_longer(cols = c(total_reads, Reads)) %>% 
  ggplot(
    aes(x = PlantID, y = p, colour = type)
  ) + 
    geom_point() + 
  scale_y_continuous(limits = c(50,100)) +
  geom_boxplot() #+
    #labs(subtitle = "Rarefied 70137 reads") +
    facet_wrap(~PlantID, scales = "free_y") 
  
```


##  Calculate alpha diversity indexes - what do they tell. What is the importance of rarefaction 

### Visualise differnet indexes

```{r}

merged_alpha_div_rare <- 
  rbind(
  d_alignment_raw %>% amp_subset_samples(minreads = 70137, rarefy = 70137) %>% 
      amp_alpha_diversity() %>% mutate(type = "alignment") %>%
      left_join(summary_reads_align %>% select(barcode, total_filtered_reads) %>% 
                  rename("SampleID" = "barcode", 
                         "total_reads" = "total_filtered_reads")),
  d_ASV %>% amp_subset_samples(minreads = 70137, rarefy = 70137) %>% 
    amp_alpha_diversity() %>% mutate(type = "ASV") %>%
      left_join(summary_reads_ASV))

merged_alpha_div %>% 
    filter(!str_detect(PlantID,  "PCR")) %>% 
  pivot_longer(cols = c(invSimpson , uniqueOTUs,  Shannon)) %>% 
  ggplot(
    aes(x = PlantID, y = value, colour = type)
  ) + 
    geom_point() + 
  geom_boxplot() +
  labs(subtitle = "Non rarefied") +
    facet_wrap(~name, scales = "free_y") +
merged_alpha_div_rare %>% 
  select(-Simpson) %>% 
    filter(!str_detect(PlantID,  "PCR")) %>% 
  pivot_longer(cols = c(invSimpson , uniqueOTUs,  Shannon)) %>% 
  ggplot(
    aes(x = PlantID, y = value, colour = type)
  ) + 
    geom_point() + 
  geom_boxplot() +
    labs(subtitle = "Rarefied 70137 reads") +
    facet_wrap(~name, scales = "free_y") + 
  plot_layout(nrow = 2)


```

### Octave plots

Generates an octave plot to assess alpha diversity. An octave plot is a histogram of the number of taxa observed by bins of read counts, where the bin ranges increase exponentially, see details.

```{r}

amp_octave(d_ASV, 
           group_by = "PlantID",
            tax_aggregate = "Species" 
           )

amp_octave(d_alignment_raw, 
                      group_by = "PlantID", 
           tax_aggregate = "Species" ,
           #tax_aggregate = "Genus"
           )


```



# How many distinct ASVs/Species/Genus/.. are present for each dataset - how many are shared? What is the effect of adding an abundance cutoff? 


```{r}
long_data_merged <- rbind(
  amp_export_long(d_alignment_norm) %>% mutate(type = "alignment"),
  amp_export_long(amp_subset_samples(d_ASV, normalise = T)) %>% mutate(type = "ASV")) %>% 
  as_tibble() %>% 
    filter(!str_detect(PlantID,  "PCR"))


```


## Number of taxa (with abundance filte)

### No filter

```{r}
long_data_merged %>% 
  filter(count > 0) %>%  ## <---  CHANGE FILTER
  group_by(type) %>% 
  summarise(
    ASVs = n_distinct(OTU), 
    Species = n_distinct(Species), 
    Genera = n_distinct(Genus), 
    Families = n_distinct(Family), 
    Orders = n_distinct(Order), 
    Phyla = n_distinct(Phylum)
  ) %>% 
  pivot_longer(cols = -c(type)) %>% 
  mutate(name = factor(name, levels = c("ASVs", "Species", "Genera", "Families", "Orders", "Phyla"))) %>% 
  ggplot(aes(x = name, y = value, fill = type)) + 
  geom_col(position = "dodge") +
  facet_wrap(~name, scales = "free", nrow = 1)


```

### 0.01

```{r}
long_data_merged %>% 
  filter(count > 0.01) %>%  ## <---  CHANGE FILTER
  group_by(type) %>% 
  summarise(
    ASVs = n_distinct(OTU), 
    Species = n_distinct(Species), 
    Genera = n_distinct(Genus), 
    Families = n_distinct(Family), 
    Orders = n_distinct(Order), 
    Phyla = n_distinct(Phylum)
  ) %>% 
  pivot_longer(cols = -c(type)) %>% 
  mutate(name = factor(name, levels = c("ASVs", "Species", "Genera", "Families", "Orders", "Phyla"))) %>% 
  ggplot(aes(x = name, y = value, fill = type)) + 
  geom_col(position = "dodge") +
  facet_wrap(~name, scales = "free", nrow = 1)


```

### 0.1

```{r}
long_data_merged %>% 
  filter(count > 0.1) %>%  ## <---  CHANGE FILTER
  group_by(type) %>% 
  summarise(
    ASVs = n_distinct(OTU), 
    Species = n_distinct(Species), 
    Genera = n_distinct(Genus), 
    Families = n_distinct(Family), 
    Orders = n_distinct(Order), 
    Phyla = n_distinct(Phylum)
  ) %>% 
  pivot_longer(cols = -c(type)) %>% 
  mutate(name = factor(name, levels = c("ASVs", "Species", "Genera", "Families", "Orders", "Phyla"))) %>% 
  ggplot(aes(x = name, y = value, fill = type)) + 
  geom_col(position = "dodge") +
  facet_wrap(~name, scales = "free", nrow = 1)


```

## Number of shared taxa (with abundance filte)

```{r}

shared_unique_taxa <- long_data_merged %>% 
  filter(count > 0.01) %>%                 ## <---  CHANGE FILTER
  select(type, OTU, Species, Genus, Family, Order, Phylum) %>% 
  pivot_longer(
    cols = c(OTU, Species, Genus, Family, Order, Phylum),
    names_to = "tax_level",
    values_to = "taxon"
  ) %>% 
  filter(!is.na(taxon), taxon != "") %>% 
  distinct(tax_level, taxon, type) %>% 
  group_by(tax_level, taxon) %>% 
  mutate(
    n_types = n_distinct(type),
    status = if_else(n_types == 1, "Unique", "Shared")
  ) %>% 
  ungroup()


shared_unique_taxa %>% 
  count(type, tax_level, status, name = "n_taxa") %>% 
  mutate(
    tax_level = factor(
      tax_level,
      levels = c("OTU", "Species", "Genus", "Family", "Order", "Phylum"),
      labels = c("ASVs", "Species", "Genera", "Families", "Orders", "Phyla")
    )
  ) %>% 
  ggplot(aes(x = status, y = n_taxa, fill = type)) +
  geom_col(position = "dodge") +
  facet_wrap(~tax_level, scales = "free_y", nrow = 2) +
  labs(
    x = NULL,
    y = "Number of taxa",
    fill = "Sample type"
  ) +
  theme_bw()
```


# What are the most abundant taxa - does if differ depending on taxonomic level?


## Species
```{r}

amp_heatmap(d_ASV %>% amp_subset_samples(!str_detect(PlantID,  "PCR")),
            group_by = "PlantID", 
            normalise = T,
            tax_show = 15,
            tax_aggregate = "Species")+
      labs(subtitle = "ASV-based") +   theme(axis.text.y = element_text(size = 10)) +
  amp_heatmap(d_alignment_norm %>% amp_subset_samples(!str_detect(PlantID,  "PCR")), 
                 group_by = "PlantID", 
              normalise = F,
              tax_show = 15,
              tax_aggregate = "Species") +
        labs(subtitle = "alignment-based")  + theme(axis.text.y = element_text(size = 10))

```

## Genus
```{r}

amp_heatmap(d_ASV %>% amp_subset_samples(!str_detect(PlantID,  "PCR")),
            group_by = "PlantID", 
            normalise = T,
            tax_show = 15,
            tax_aggregate = "Genus")+
      labs(subtitle = "ASV-based") +   theme(axis.text.y = element_text(size = 10)) +
  amp_heatmap(d_alignment_norm %>% amp_subset_samples(!str_detect(PlantID,  "PCR")), 
                 group_by = "PlantID", 
              normalise = F,
              tax_show = 15,
              tax_aggregate = "Genus") +
        labs(subtitle = "alignment-based")  + theme(axis.text.y = element_text(size = 10))

```

## Family

```{r}

amp_heatmap(d_ASV %>% amp_subset_samples(!str_detect(PlantID,  "PCR")),
            group_by = "PlantID", 
            normalise = T,
            tax_show = 15,
            tax_aggregate = "Family")+
      labs(subtitle = "ASV-based") +   theme(axis.text.y = element_text(size = 10)) +
  amp_heatmap(d_alignment_norm %>% amp_subset_samples(!str_detect(PlantID,  "PCR")), 
                 group_by = "PlantID", 
              normalise = F,
              tax_show = 15,
              tax_aggregate = "Family") +
        labs(subtitle = "alignment-based")  + theme(axis.text.y = element_text(size = 10))

```



# Consider how to merge dataset to visualize betadiversity

--> no match between the ASVs 
--> Aggregate at higher taxonomic rank e.g. Species or Genus

*Install the vegan package*

## Species

```{r}

make_df <- long_data_merged %>% 
  group_by(Species, Barcode, type, PlantID) %>% 
  summarise(rel_abun = sum(count)) %>%
  ungroup() %>% 
  mutate(Species = if_else(Species == "", "Unclassified", Species)) %>% 
  filter(Species != "Unclassified")  %>%     ## Remove unclassified taxa 
  select(Barcode, type, PlantID,Species, rel_abun) %>% 
  mutate(SampleID = paste0(type, "_", Barcode)) %>% relocate(SampleID) %>% 
  filter(rel_abun > 0.001) 

# Format as matrix
species_mat <- make_df %>% 
  pivot_wider(
    names_from = Species,
    values_from = rel_abun,
    values_fill = 0
  )

# Metadata
metadata <- species_mat %>%
  select(SampleID, type, PlantID)

comm <- species_mat %>%
  select(-SampleID, -Barcode, -type, -PlantID) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.data.frame()

rownames(comm) <- metadata$SampleID

# Calculate distance matrix
bc_dist <- vegan::vegdist(comm, method = "bray")


# PCoA
pcoa <- cmdscale(bc_dist, eig = TRUE, k = 2)

pcoa_df <- data.frame(
  SampleID = rownames(comm),
  Axis1 = pcoa$points[,1],
  Axis2 = pcoa$points[,2]
) %>%
  left_join(metadata, by = "SampleID")

# Variance explained
var_exp <- round(100 * pcoa$eig / sum(pcoa$eig[pcoa$eig > 0]), 1)

# Plot
ggplot(
  pcoa_df,
  aes(
    x = Axis1,
    y = Axis2,
    color = PlantID,
    shape = type
  )
) +
  geom_point(size = 3) +
  labs(
    x = paste0("PCoA1 (", var_exp[1], "%)"),
    y = paste0("PCoA2 (", var_exp[2], "%)")
  ) +
  theme_bw()

```

## Species - one plant

```{r}

make_df <- long_data_merged %>% 
  filter(PlantID == "AAW") %>% 
  group_by(Species, Barcode, type, PlantID) %>% 
  summarise(rel_abun = sum(count)) %>%
  ungroup() %>% 
  mutate(Species = if_else(Species == "", "Unclassified", Species)) %>% 
  filter(Species != "Unclassified")  %>%     ## Remove unclassified taxa 
  select(Barcode, type, PlantID,Species, rel_abun) %>% 
  mutate(SampleID = paste0(type, "_", Barcode)) %>% relocate(SampleID) %>% 
  filter(rel_abun > 0.01) 

# Format as matrix
species_mat <- make_df %>% 
  pivot_wider(
    names_from = Species,
    values_from = rel_abun,
    values_fill = 0
  )

# Metadata
metadata <- species_mat %>%
  select(SampleID, type, PlantID)

comm <- species_mat %>%
  select(-SampleID, -Barcode, -type, -PlantID) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.data.frame()

rownames(comm) <- metadata$SampleID

# Calculate distance matrix
bc_dist <- vegan::vegdist(comm, method = "bray")


# PCoA
pcoa <- cmdscale(bc_dist, eig = TRUE, k = 2)

pcoa_df <- data.frame(
  SampleID = rownames(comm),
  Axis1 = pcoa$points[,1],
  Axis2 = pcoa$points[,2]
) %>%
  left_join(metadata, by = "SampleID")

# Variance explained
var_exp <- round(100 * pcoa$eig / sum(pcoa$eig[pcoa$eig > 0]), 1)

# Plot
ggplot(
  pcoa_df,
  aes(
    x = Axis1,
    y = Axis2,
    color = PlantID,
    shape = type
  )
) +
  geom_point(size = 3) +
  labs(
    x = paste0("PCoA1 (", var_exp[1], "%)"),
    y = paste0("PCoA2 (", var_exp[2], "%)")
  ) +
  theme_bw()

```

## Genus

```{r}

make_df <- long_data_merged %>% 
  group_by(Genus, Barcode, type, PlantID) %>% 
  summarise(rel_abun = sum(count)) %>%
  ungroup() %>% 
  mutate(Genus = if_else(Genus == "", "Unclassified", Genus)) %>% 
  filter(Genus != "Unclassified")  %>%     ## Remove unclassified taxa 
  select(Barcode, type, PlantID,Genus, rel_abun) %>% 
  mutate(SampleID = paste0(type, "_", Barcode)) %>% relocate(SampleID) %>% 
  filter(rel_abun > 0.01) 

# Format as matrix
species_mat <- make_df %>% 
  pivot_wider(
    names_from = Genus,
    values_from = rel_abun,
    values_fill = 0
  )

# Metadata
metadata <- species_mat %>%
  select(SampleID, type, PlantID)

comm <- species_mat %>%
  select(-SampleID, -Barcode, -type, -PlantID) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.data.frame()

rownames(comm) <- metadata$SampleID

# Calculate distance matrix
bc_dist <- vegan::vegdist(comm, method = "bray")


# PCoA
pcoa <- cmdscale(bc_dist, eig = TRUE, k = 2)

pcoa_df <- data.frame(
  SampleID = rownames(comm),
  Axis1 = pcoa$points[,1],
  Axis2 = pcoa$points[,2]
) %>%
  left_join(metadata, by = "SampleID")

# Variance explained
var_exp <- round(100 * pcoa$eig / sum(pcoa$eig[pcoa$eig > 0]), 1)

# Plot
ggplot(
  pcoa_df,
  aes(
    x = Axis1,
    y = Axis2,
    color = PlantID,
    shape = type
  )
) +
  geom_point(size = 3) +
  labs(
    x = paste0("PCoA1 (", var_exp[1], "%)"),
    y = paste0("PCoA2 (", var_exp[2], "%)")
  ) +
  theme_bw()

```

## Genus - one plant

```{r}

make_df <- long_data_merged %>% 
    filter(PlantID == "AAW") %>% 
  group_by(Genus, Barcode, type, PlantID) %>% 
  summarise(rel_abun = sum(count)) %>%
  ungroup() %>% 
  mutate(Genus = if_else(Genus == "", "Unclassified", Genus)) %>% 
  filter(Genus != "Unclassified")  %>%     ## Remove unclassified taxa 
  select(Barcode, type, PlantID,Genus, rel_abun) %>% 
  mutate(SampleID = paste0(type, "_", Barcode)) %>% relocate(SampleID) %>% 
  filter(rel_abun > 0.01) 

# Format as matrix
species_mat <- make_df %>% 
  pivot_wider(
    names_from = Genus,
    values_from = rel_abun,
    values_fill = 0
  )

# Metadata
metadata <- species_mat %>%
  select(SampleID, type, PlantID)

comm <- species_mat %>%
  select(-SampleID, -Barcode, -type, -PlantID) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.data.frame()

rownames(comm) <- metadata$SampleID

# Calculate distance matrix
bc_dist <- vegan::vegdist(comm, method = "bray")


# PCoA
pcoa <- cmdscale(bc_dist, eig = TRUE, k = 2)

pcoa_df <- data.frame(
  SampleID = rownames(comm),
  Axis1 = pcoa$points[,1],
  Axis2 = pcoa$points[,2]
) %>%
  left_join(metadata, by = "SampleID")

# Variance explained
var_exp <- round(100 * pcoa$eig / sum(pcoa$eig[pcoa$eig > 0]), 1)

# Plot
ggplot(
  pcoa_df,
  aes(
    x = Axis1,
    y = Axis2,
    color = PlantID,
    shape = type
  )
) +
  geom_point(size = 3) +
  labs(
    x = paste0("PCoA1 (", var_exp[1], "%)"),
    y = paste0("PCoA2 (", var_exp[2], "%)")
  ) +
  theme_bw()

```


# At which taxonomic level would you trust the datasets? Why?


