---
title: "MiDAS course: Hands on - part 2"
author: "Marie Riisgaard-Jensen"
date: "`r format(Sys.time(), '%Y-%m-%d')`, Aalborg, Denmark"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Questions for the MiDAS course - ampvis hands on part 2. 


## Load the packages

```{r, warning = F, message=F}
library(ampvis2)
library(tidyverse)
library(patchwork) ## Combine different plot using '+'

```

## Set your working directory - only if you did not open script from the downloaded folder
```{r, eval = FALSE}
#setwd("") 
```


# Details about the pipelines

**ASV based: **
Pipeline: https://github.com/KasperSkytte/snakemake_usearch 

**Alignment based (95% identity)**
Pipeline: https://github.com/cmc-aau/nanopore_16Samp 

## Load the data into an ampvis object

- Rember to adjust paths

```{r}

setwd("...")


metadata <- read.csv("metadata.csv") %>%
  mutate(SampleID = Barcode) %>% relocate(SampleID) 


# Alignment based (minimap min identity 95%)
## Raw read (in counts)
d_alignment_raw <- amp_load(otutable = "Alignment-based/alignment_otutable_mappedreads.tsv",
              metadata = metadata) 
## Normalised data (in percentage: with relative abundances are calculated taking non-mapped reads into account)
d_alignment_norm <- amp_load(otutable = "Alignment-based/alignment_otutable_normalised.tsv",
              metadata = metadata) 

## Summary file 
summary_reads_align <- read.csv("Alignment-based/alignment_summary.txt") 



# ASV based 
d_ASV <- amp_load(otutable = "ASV-based/ASV_abund_table.tsv",
              taxonomy = "ASV-based/ASV_zOTUs.sintax",
              metadata = metadata
              ) 
# Total reads 
summary_reads_ASV <- read.csv("ASV-based/ASV_totalreads.csv") 

```


# Questions: 

Explain and/or make plots i order to answer the questions

- What are the main difference between the datasets? 
- How are unclassified reads treated?
- How many reads are unmapped in the two datasets? What do these represent? 
- Calculate alpha diversity indexes - what do they tell. What is the importance of rarefaction 
- How many distinct ASVs/Species/Genus/.. are present for each dataset - how many are shared? What is the effect of adding an abundance cutoff? 
- What are the most abundant taxa - does if differ depending on taxonomic level?
- Consider how to merge dataset to visualize betadiversity
- At which taxonomic level would you trust the datasets? Why?  





